---
title: New Year, Same Moons
author: Jacquie Tran
date: '2025-03-27'
slug: []
categories:
  - On-field analyses
tags:
  - cricket
  - r stats
  - sport science
  - sports analytics
---

# Introduction

The Australian Women's Cricket team have wrapped up their very long international season for 2024/2025 with another T20 series sweep, this time winning 3-0 over the White Ferns.

And who was at the centre of the action? Beth Mooney piling on the runs. Naturally.

New year, same run scoring machine.

Cricket Australia reported that, [in 2025, Moons has scored "575 runs at an average of 72"](https://www.cricket.com.au/news/4239115/georgia-voll-third-t20i-australia-new-zealand-wellington-opener-beth-mooney-australia-a-innings?tags=9037). Let's dig into the details of this scoring spree from the ever-metronomic Mooney.

# Set up and retrieve data

We'll use ball-by-ball data provided by Cricsheet.org, across the matches played by Australia in 2025 (T20Is, ODIs, and Test):

```{r setup, message = FALSE}

# Load libraries ---------------------------------------------------------------

library(cricketdata) # for retrieving cricket data from ESPNCricinfo and Cricsheet
library(dplyr)       # for tidying data
library(gt)          # for visualising data tables
library(showtext)    # for plotting
library(ggplot2)     # for plotting

```

```{r retrieve_data, message = FALSE, cache = TRUE}

# Retrieve ball-by-ball data ---------------------------------------------------

# Women's T20Is
t20s_bbb <- fetch_cricsheet(
  type = "bbb",
  competition = "t20s",
  gender = "female")

# Women's ODIs
odis_bbb <- fetch_cricsheet(
  type = "bbb",
  competition = "odis",
  gender = "female")

# Women's Tests
tests_bbb <- fetch_cricsheet(
  type = "bbb",
  competition = "tests",
  gender = "female")

```

# Tidy the data

Let's get the retrieved data sets into more focused subsets, focusing on balls where Beth Mooney was on strike:

``` {r subset_and_bind_data}

# Filter to matches played in 2025 and Beth Mooney on strike -------------------

# T20s
t20s_bbb_2025_mooney <- t20s_bbb %>%
  filter(
    start_date > "2025-01-01" & striker == "BL Mooney") %>%
  mutate(
    competition = "T20I",
    start_date = as.character(start_date) # For binding purposes in next step
    )

# ODIs
odis_bbb_2025_mooney <- odis_bbb %>%
  filter(
    start_date > "2025-01-01" & striker == "BL Mooney") %>%
  mutate(competition = "ODI")

# Test
tests_bbb_2025_mooney <- tests_bbb %>%
  filter(
    start_date > "2025-01-01" & striker == "BL Mooney") %>%
  mutate(competition = "Test")

# Bind subsets into one data frame ---------------------------------------------

mooney_in_2025 <- bind_rows(
  t20s_bbb_2025_mooney, odis_bbb_2025_mooney, tests_bbb_2025_mooney)

```

# A glance off the pads: Batting stats by format

```{r calculate_summary_stats}

mooney_in_2025_summary <- mooney_in_2025 %>%
  mutate(
    match_innings_id = paste0(match_id, innings, sep = "_"),
    is_ball_faced = case_when(
      wides == 1 ~ 0, # Wides do not count as a ball faced by the batter
      TRUE       ~ 1),
    wicket_counting_helper = case_when(
      player_dismissed == striker ~ 1,
      TRUE                        ~ 0)) %>%
  group_by(competition) %>%
  summarise(
    runs_off_bat = sum(runs_off_bat),
    balls_faced = sum(is_ball_faced),
    innings_batted = length(unique(match_innings_id)),
    wickets_lost = sum(wicket_counting_helper),
    .groups = "drop"
  ) %>%
  mutate(
    average_in_2025 = round(runs_off_bat / wickets_lost, 2),
    strike_rate = round(runs_off_bat / balls_faced * 100, 2))

```

```{r tabulate_summary_stats}

mooney_in_2025_summary_table <- mooney_in_2025_summary %>%
  gt(rowname_col = "competition") %>%
  tab_header(
    title = md("**Piling on the runs to see out the 2024/25 season**"),
    subtitle = "Mooney's batting stats from internationals - January to March 2025") %>%
  cols_label(
    runs_off_bat = "Runs",
    balls_faced = "Balls faced",
    innings_batted = "Innings",
    wickets_lost = "Wickets lost",
    average_in_2025 = "Avg. in 2025",
    strike_rate = "Strike rate") %>%
  tab_style(
    style = cell_fill(color = "#9AD4D6"),
    locations = cells_body(
      rows = competition == "T20I")) %>%
  tab_source_note(
    source_note = "Data source: Cricsheet.org")

```

When we look at Mooney's batting separated by format, we can see her dominance in recent T20s more clearly, where she averaged 95 runs with a strike rate of 154, accumulating nearly 400 runs across only 6 innings!

```{r echo = FALSE}

mooney_in_2025_summary_table

```

# When does Moons hit the gas?

I was interested in whether there has been a pattern in when Moons tends to accelerate with her batting. Focusing on T20I innings, I decided to plot her cumulative runs scored against cumulative balls faced in each innings:

```{r analyse_t20_scoring_rate}

t20s_bbb_2025_mooney_tidy <- t20s_bbb_2025_mooney %>%
  mutate(
    is_ball_faced = case_when(
      wides == 1 ~ 0, # Wides do not count as a ball faced by the batter
      TRUE       ~ 1)) %>%
  # Filter out wides
  filter(is_ball_faced == 1) %>%
  group_by(match_id) %>%
  mutate(
    balls_faced_cumulative = cumsum(is_ball_faced),
    runs_scored_cumulative = cumsum(runs_off_bat)) %>%
  ungroup() %>%
  mutate(
    balls_faced_grouping = case_when(
      balls_faced_cumulative <= 10   ~ "1-10",
      balls_faced_cumulative > 10 &
        balls_faced_cumulative <= 20 ~ "11-20",
      balls_faced_cumulative > 20 &
        balls_faced_cumulative <= 30 ~ "21-30",
      balls_faced_cumulative > 30 &
        balls_faced_cumulative <= 40 ~ "31-40",
      balls_faced_cumulative > 40 ~ "41+"))

```

```{r plot_t20_scoring_rate, message = FALSE}

# Prep to plot -----------------------------------------------------------------

# Turn on {showtext} functionality
showtext_auto()

# Add Google Fonts
font_add_google("IBM Plex Sans Condensed", "plexc")

# Build plot -------------------------------------------------------------------

plot_t20_scoring_rate <- ggplot() +
  geom_point(
    data = t20s_bbb_2025_mooney_tidy,
    aes(
      x = balls_faced_cumulative, y = runs_scored_cumulative,
      group = match_id),
    size = 3, shape = 16, alpha = 0.2) +
  geom_smooth(
    data = t20s_bbb_2025_mooney_tidy,
    aes(
      x = balls_faced_cumulative, y = runs_scored_cumulative,
      color = factor(balls_faced_grouping)),
    method = "lm", se = F, linewidth = 2) +
  labs(
    title = "When does Beth Mooney hit the gas? From start to finish!",
    subtitle = "Beth Mooney's batting in T20Is - January to March 2025",
    x = "Balls faced (cumulative)",
    y = "Runs scored (cumulative)",
    caption = "Data source: Cricsheet.org") +
  scale_colour_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(family = "plexc"),
    plot.title = element_text(size = rel(3), face = "bold"),
    plot.subtitle = element_text(size = rel(2.5)),
    plot.caption = element_text(
      size = rel(2), margin = margin(t = 10)),
    axis.title = element_text(size = rel(2.5)),
    axis.text = element_text(size = rel(2))
  )

```

I genuinely laughed out loud when I generated this plot:

```{r echo = FALSE, message = FALSE}

plot_t20_scoring_rate

# Turn off {showtext} functionality
showtext_auto(FALSE)

```

I plotted separate linear trendlines for balls 1-10, 11-20, 21-30, 31-40, and 40+. The slope of each trendline is pretty much the same! Which tells us that Moons' nickname of ["Ms Consistent"](https://www.cricket.com.au/news/4235969/beth-mooney-200-international-matches-australia-eden-park-auckland-first-t20-new-zealand-milestone-runs) is well-deserved in many ways - her rate of scoring in recent T20Is has been remarkably even right through an innings.

There's only 6 T20I innings represented here, so it'd be worth including more innings to see if a more defined "acceleration phase" is evident or not. I'd also be interested to see what this plot looks like for other batters.

