---
title: How much do you get paid?
author: Jacquie Tran
date: '2017-12-26'
slug: how-much-do-you-get-paid
categories:
  - Random projects
tags:
  - sport science
  - strength and conditioning
  - soccer
  - football
  - salaries
draft: yes
---

It's a touchy subject for many. Yet this very question was posed by [Aiden Oakley](www.twitter.com/ajokaley) (**@ajoakley** on Twitter), and willingly answered by over 100 sport scientists and strength & conditioning coaches who currently work for football (soccer) clubs in the UK.

Why?

This job advert says it all...

<center><blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">This is shocking! &quot;Dear Candidate, please spend £27,000 on BSc, then £1000 on ASCC, to be paid less than half of this. Oh, and the players earn your salary in one week.&quot; 🤔🤔🤔 <a href="https://t.co/6efKtAaaQl">pic.twitter.com/6efKtAaaQl</a></p>&mdash; Rob Chave (@Robchave7) <a href="https://twitter.com/Robchave7/status/941234320034824192?ref_src=twsrc%5Etfw">December 14, 2017</a></blockquote></center>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

Blackpool FC was roundly criticised for advertising such a position for a measly £12,000 per annum. The UK Strength & Conditioning Association then withdrew this job posting from their website. Regrettably, this is not a one-off. Blackpool happened to rightly draw the ire of the Twitter crowd on this occasion, but many of us working in sport have seen too many job ads just like this one.

So what is the current state of play when it comes to salaries in UK soccer?

Aiden set out to explore this question by surveying a sample of sport scientists and strength and conditioning coaches via Twitter:

<center><blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Off the back of Blackpoolgate are any S&amp;C&#39;s/Sport Scientists willing to fill in a short questionnaire on salaries in football. Open to Premier League - League 2 and Scottish Premiership clubs,1st team &amp; Academy.Will be shared on Twitter anonymously. Plz RT <a href="https://t.co/Gcyqiu0MP2">https://t.co/Gcyqiu0MP2</a></p>&mdash; Aiden Oakley (@ajoakley) <a href="https://twitter.com/ajoakley/status/942004931279536128?ref_src=twsrc%5Etfw">December 16, 2017</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
</center>

After a fortnight, 100+ responses were received. I put my hand up to help with visualising the data.

## Getting started
Firstly, let's load the cleaned data in and load the necessary libraries. _Note: Some variables were omitted during the data cleaning process to protect the anonymity of respondents._

``` {r globalOptions, echo = FALSE}

knitr::opts_chunk$set(fig.width = 7,
											fig.height = 5,
											fig.align = 'center',
											echo = TRUE,
											warning = FALSE,
											message = FALSE)

```

``` {r setup}

library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)

# Read the data into R
dataSource <- "https://raw.githubusercontent.com/jacquietran/soccerSalariesDataPublic/master/soccerSalariesDataPublic.csv"
soccerSalariesData <- read_delim(dataSource, delim = ",", col_names = TRUE, col_types = NULL)

```

Now that we have the cleaned data loaded in, we can begin to examine it. The data set has 16 variables and 102 responses. For pedants like me, you can scroll to the bottom of this post to see the data dictionary.

``` {r checkData}

# Examine first few observations of the data set
head(soccerSalariesData)

```

## Salaries by league
Let's start with a simple plot to answer a simple question: **what are the reported salaries of performance support staff working in UK soccer clubs when categorised by league?** (It's worth noting here that this refers to the league in which the club's first team plays, so it may not be the league that the respondent may predominantly work within, e.g., if they support the Academy programs.)

After a quick exploration when I first started working with this data set, I noticed that salaries varied considerably between respondents. To capture this, I decided to build box plots (to represent 'typical' salaries within a particular category), overlaid with a scatter plot where each dot represents one respondent. I also chose to present the the x axis categories such that the leagues progress from higher tier to lower tier as your gaze moves left to right.

``` {r plotByLeague}

# Set factor levels
soccerSalariesData$leagueOfFirstTeam <- factor(
	soccerSalariesData$leagueOfFirstTeam,
	levels = c(
		"Premier League", "Scottish Premiership", "Championship", "League 1", "League 2"))

# Create object to store axis labels with line breaks
labelsLeague <- c(
	"Premier\nLeague", "Scottish\nPremiership", "Championship", "League 1", "League 2")

# Create a list to store common theme options for use across plots
plotFeatures <- theme(
	plot.title = element_text(size = 16),
	plot.subtitle = element_text(size = 14),
	axis.title.x = element_blank(),
  axis.text.x = element_text(size = 11),
	axis.title.y = element_text(size = 16),
	axis.text.y = element_text(size = 11),
	legend.position = "none")

p <- ggplot(soccerSalariesData, aes(x = leagueOfFirstTeam, y = salaryGBP))
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = leagueOfFirstTeam))
p <- p + scale_y_continuous(
	limits = c(0, 60000),
	breaks = seq(0, 60000, by = 10000))
p <- p + scale_x_discrete(labels = labelsLeague)
p <- p + labs(
	title = "Soccer support staff salaries in the UK",
	subtitle = "Categorised by league that the first team plays in",
	y = "Salary (GBP)")
p <- p + plotFeatures
salaryByLeague <- p

# Display the plot
salaryByLeague

```

What about median reported salaries?

``` {r medianSalariesByLeague}

# Calculate median salary by league in which first team plays
medianSalariesByLeague <- soccerSalariesData %>%
	group_by(leagueOfFirstTeam) %>%
	summarise(median = median(salaryGBP, na.rm = TRUE))

# Rename variables for aesthetic purposes
colnames(medianSalariesByLeague) <- c("League in which first team plays", "Median salary (GBP, £)")

# Display the table
medianSalariesByLeague %>%
	kable("html") %>%
	kable_styling()

```

What do these results indicate?

**1. Across the board, typical salaries are low.**

- Given that 80 respondents were qualified at Masters or PhD level, these figures seem low. Such individuals might be competitive for university lecturer positions. [Coventry University](https://staffrecruitment.coventry.ac.uk/tlive_webrecruitment/wrd/run/ETREC107GF.open?VACANCY_ID=954467Gmbh&WVID=1861420Izv&LANG=USA&utm_source=Indeed&utm_medium=organic&utm_campaign=Indeed) is advertising for an Assistant Lecturer in Sport and Exercise Science with a stated salary range of ~£26,000--£35,000 per annum. Meanwhile, a Lecturer / Senior Lecturer in Research Methods role at [Nottingham Trent](https://vacancies.ntu.ac.uk/displayjob.aspx?jobid=4698) will pay somewhere in the range of ~£33,000--£48,000 per annum.
- FYI the [UK's National Living Wage](https://www.gov.uk/national-minimum-wage-rates) for individuals aged 25 and over is £7.50 per hour (as from April 2017). With some 'back of the napkin' math, this works out to a full-time salary of approximately £12,600 per annum - calculated as £7.50 per hour x (35 hours / week) x (48 weeks / year).
- For the Australians among us (_puts hand up_), the typical salary for a performance support staff member working for a Scottish Premiership team—£26,000—works out to roughly $45,000 AUD (based on the exchange rate on 28/12/2017). Level A academics are paid approximately $64,000--$86,000 and Level B academics are paid approximately $91,000--$18,000 (according to [Deakin's pay scales](https://www.deakin.edu.au/__data/assets/pdf_file/0018/41472/salary-rates-academic.pdf) from April 2016).
 
**2. Performance support staff working for Scottish Premiership teams appear to be paid slghtly better than their counterparts in the Premier League.** Having said that, it's quite possible that this difference may be due to the small sub-sample of respondents in the Scottish Premiership category (n = 6).

**3. There is considerable variation within most leagues.**

- The '£0' salaries have been left in the data set as a handful of (unpaid!!!) interns completed the questionnaire. Even if you were to remove these data points, though, sizeable variation remains. Also, the median is robust to outliers, so typical salaries do not change or change by a negligible amount when interns are excluded from the data set.
- This is partially due to some respondents working part-time in their roles. It was possible to identify whether some individuals were working part-time, based on the data entered by some participants, but the questionnaire itself did not explicitly ask this question.

# Salaries by league and competitive level
We can drill down further, plotting the data by league _and_ by competitive level; that is, whether respondents work for their club's first team or academy.

``` {r plotByLeagueAndCompLevel, fig.width = 10, fig.height = 5}

p <- ggplot(soccerSalariesData, aes(x = leagueOfFirstTeam, y = salaryGBP))
p <- p + facet_wrap(~levelOfWork)
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = leagueOfFirstTeam))
p <- p + scale_y_continuous(
	limits = c(0, 60000),
	breaks = seq(0, 60000, by = 10000))
p <- p + scale_x_discrete(labels = labelsLeague)
p <- p + labs(
	title = "Soccer support staff salaries in the UK",
	subtitle = "First Team vs. Academy",
	y = "Salary (GBP)")
p <- p + plotFeatures
salaryByLeagueAndCompLevel <- p

# Display the plot
salaryByLeagueAndCompLevel

```

Meanwhile, median salaries now look like this:

``` {r medianSalariesByLeagueAndCompLevel}

# Calculate median salaries by league and competitive level
medianSalariesByLeagueAndCompLevel <- soccerSalariesData %>%
	group_by(leagueOfFirstTeam, levelOfWork) %>%
	summarise(median = median(salaryGBP, na.rm = TRUE))

# Order by $levelOfWork, then $leagueOfFirstTeam
medianSalariesByLeagueAndCompLevel <- medianSalariesByLeagueAndCompLevel[order(
	medianSalariesByLeagueAndCompLevel$levelOfWork,
	medianSalariesByLeagueAndCompLevel$leagueOfFirstTeam), ]

# Revalue duplicates in $levelOfWork for aesthetic purposes
medianSalariesByLeagueAndCompLevel$levelOfWork <- ifelse(
	duplicated(medianSalariesByLeagueAndCompLevel$levelOfWork),
	"", medianSalariesByLeagueAndCompLevel$levelOfWork)

# Rename variables for aesthetic purposes
colnames(medianSalariesByLeagueAndCompLevel) <- c(
	"League in which first team plays", "Competitive level", "Median salary (GBP, £)")

# Display the table
medianSalariesByLeagueAndCompLevel %>%
	kable("html") %>%
	kable_styling()

```

Categorising by league and what I've termed 'competitive level' (first team vs. academy) helps to better visualise the salary gradient across the different leagues. This segmentation of the data gives us a better look at salary breakdowns, but also divides the relatively small sample into even smaller sub-groups. For instance, there were only 2 respondents who work for a Scottish Premiership first team, and only 3 respondents who work for a League 2 academy. Anyway, the data is the data, so let's press on while keeping in mind that _the results seen here may not hold with a larger sample and more representation within each sub-group._

Caveats notwithstanding, the results show that:

**1. The typical salary of Premier League first team support staff (£40,000) is higher than for those who work for Scottish Premiership first teams (£31,250).**

**2. The differences between leagues are stark in terms of first team staff salaries, while the gradient is not so steep when it comes to academy staff salaries.**

## Salaries by position
What do we see when we categorise by job 'seniority' (and exclude responses from unpaid interns)?

``` {r plotByPosition}

# Subset of data frame to exclude position == NA | Intern
soccerSalariesData2 <- soccerSalariesData %>%
	filter(position != "Intern" & !is.na(position))
# Set factor levels
soccerSalariesData2$position <- factor(soccerSalariesData2$position,
																			 levels = c("Head", "Lead", "Coach"))

p <- ggplot(soccerSalariesData2, aes(x = position, y = salaryGBP))
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = position))
p <- p + scale_y_continuous(limits = c(0, 60000),
														breaks = seq(0, 60000, by = 10000))
p <- p + labs(title = "Soccer support staff salaries in the UK",
							subtitle = "Categorised by position: Department head, discipline lead, or 'coach'",
							y = "Salary (GBP)")
p <- p + plotFeatures
salaryByPosition <- p

# Display the plot
salaryByPosition

```

## Salaries by league and by discipline
Let's compare how sport scientists and S&C coaches are paid in the different leagues.

``` {r plotByLeagueAndDiscipline}

# Subset of data frame to exclude dept == NA
soccerSalariesData3 <- soccerSalariesData %>%
	filter(!is.na(dept))
soccerSalariesData3$dept <- recode(soccerSalariesData3$dept,
																	 `Physical Performance` = "S&C")
# Set factor levels
soccerSalariesData3$dept <- factor(soccerSalariesData3$dept,
																	 levels = c("S&C", "Sport Science", "Both"))

p <- ggplot(soccerSalariesData3, aes(x = leagueOfFirstTeam, y = salaryGBP))
p <- p + facet_wrap(~dept)
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = leagueOfFirstTeam))
p <- p + scale_y_continuous(limits = c(0, 60000),
														breaks = seq(0, 60000, by = 10000))
p <- p + scale_x_discrete(labels = labelsLeague)
p <- p + labs(title = "Soccer support staff salaries in the UK",
							subtitle = "Categorised by discipline",
							y = "Salary (GBP)")
p <- p + plotFeatures
salaryByLeagueAndDept <- p

# Display the plot
salaryByLeagueAndDept

```

## Salaries by academy category
The FA sets the academy status of all football clubs with academy programs. Budgets are somewhat dependent on guidelines that are category specific. How does academy category affect salary of strength and conditioning coaches and sport scientists?

``` {r plotByAcademyCategory}

# Subset of data frame to exclude academyCategory == NA
soccerSalariesData4 <- soccerSalariesData %>%
	filter(!is.na(academyCategory))

# Set factor levels
soccerSalariesData4$academyCategory <- factor(soccerSalariesData4$academyCategory,
																							levels = c("1", "2", "3", "SFA Elite", "Development"))

# Create object to store axis labels with line breaks
labelsAcademyCategory <- c("England\nCategory 1", "England\nCategory 2", "England\nCategory 3",
													 "Scotland\nSFA Elite", "Scotland\nDevelopment")

p <- ggplot(soccerSalariesData4, aes(x = academyCategory, y = salaryGBP))
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = academyCategory))
p <- p + scale_y_continuous(limits = c(0, 50000),
														breaks = seq(0, 50000, by = 10000))
p <- p + scale_x_discrete(labels = labelsAcademyCategory)
p <- p + labs(title = "Soccer support staff salaries in the UK: Academy staff only",
							subtitle = "Organised by Academy Category",
							y = "Salary (GBP)")
p <- p + plotFeatures
salaryByAcademyCategory <- p

# Display the plot
salaryByAcademyCategory

```

## Salaries by highest qualification
``` {r plotByHighestQual}

# Set factor levels
soccerSalariesData$highestQualCategory <- factor(soccerSalariesData$highestQualCategory,
																								 levels = c(
																								 	"PhD", "Postgraduate", "Undergraduate", "Unclear"))

p <- ggplot(soccerSalariesData, aes(x = highestQualCategory, y = salaryGBP))
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = highestQualCategory))
p <- p + scale_y_continuous(limits = c(0, 50000),
														breaks = seq(0, 50000, by = 10000))
p <- p + labs(title = "Soccer support staff salaries in the UK",
							subtitle = "Categorised by highest qualification",
							y = "Salary (GBP)")
p <- p + plotFeatures
salaryByHighestQual <- p

# Display the plot
salaryByHighestQual

```

## Ways to improve the questionnaire
- Future questionnaires need to specifically ask participants to identify the nature of their work in terms of whether they are full-time, part-time, casual, or volunteer / intern. It would also be helpful to know, for those not in full-time roles especially, an estimate of hours worked per week. This would allow pro rata annual salaries and hourly rates to be calculated.

## Want to explore the data?
If you'd like to take a look at the data and try your hand at creating / modifying these plots, visit this [GitHub repostiory](www.github.com/jacquietran/soccerSalariesDataPublic).

Also, here's the data dictionary I promised earlier:

``` {r dataDictionary, echo = FALSE}

vars <- as.vector(names(soccerSalariesData))
varDescriptions <- c(
	"ID number for respondent",
	"Country in which the respondent's club competes",
	"League that the club's first team plays in",
	"Level of respondent's work - categorical: 1st Team, Academy",
	"Respondent's position",
	"Respondent's department / discipline",
	"Variable for identifying individuals who noted they are part-time employed at their club",
	"For those working at Academy level, the Academy category status (according to designations by the Football Association (FA) or the Scottish Football Association",
	"For those working at Academy level, whether they support athletes across all Academy phases (as per age group designations set by the FA)",
	"For those working at Academy level, whether they support athletes in the Professional Development phase (U17s and older)",
	"For those working at Academy level, whether they support athletes in the Youth Development phase (U12s to U16s, inclusive)",
	"For those working at Academy level, whether they support athletes in the Foundation Phase (U9s to U11s, inclusive)",
	"Reported salary",
	"Salary, including reported values and estimated values based on additional comments provided by respondent (e.g., salary in GBP estimated based on conversion from another currency)",
	"Respondent's highest qualification - text as entered by participant",
	"Respondent's highest qualification (categorical format)")
dataClasses <- c("Numeric",
								 "Categorical - England, Scotland",
								 "Categorical - Premier League, Scottish Premiership, Championship, League 1, League 2",
								 "Categorical: 1st Team, Academy",
								 "Categorical: Head (of department), Lead (senior staff member in a particular discipline), Coach (not Head or Lead positions), Intern, NA",
								 "Categorical: Sport Science, S&C (Strength & Conditioning), or Both",
								 "String: Known part-time staff noted as 'Part Time'",
								 "Categorical: 1, 2, 3, SFA Elite, Development",
								 "Logical: TRUE, FALSE",
								 "Logical: TRUE, FALSE",
								 "Logical: TRUE, FALSE",
								 "Logical: TRUE, FALSE",
								 "Numeric - currency, GBP (£)",
								 "Numeric - currency, (£)",
								 "String",
								 "Categorical: PhD, Postgraduate (higher than Bachelor but below PhD), Undergraduate, Unclear (based on respondent's entry)")

dataDictionary <- cbind(vars, varDescriptions, dataClasses)
colnames(dataDictionary) <- c("Variable name", "Description", "Data type")

dataDictionary %>%
	kable("html") %>%
	kable_styling()

```