---
title: How much do you get paid? Part V - Credential creep
author: Jacquie Tran
date: '2018-01-08'
slug: how-much-do-you-get-paid-part-5
categories:
  - Off-field analyses
tags:
  - football
  - salaries
  - soccer
  - sport science
  - strength and conditioning
draft: yes
---

## Introduction
This post is the fifth in a series - go on and read the **first**, **second**, **third**, and **fourth** posts for background. In this series, I will show how I created the plots that Aiden published on Twitter [here](www.twitter.com/ajoakley/status/944564698166513665) and [here](https://twitter.com/ajoakley/status/944936008285552642), alongside some commentary about the choices I made throughout the process and my own interpretations of the results.

## Getting started
Load the data and the libraries **as we've done before**.

``` {r globalOptions, echo = FALSE}

knitr::opts_chunk$set(fig.width = 7,
											fig.height = 5,
											fig.align = 'center',
											echo = TRUE,
											warning = FALSE,
											message = FALSE)

```

``` {r setup}

library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)

# Read the data into R
dataSource <- "https://raw.githubusercontent.com/jacquietran/soccerSalariesDataPublic/master/soccerSalariesDataPublic.csv"
soccerSalariesData <- read_delim(dataSource, delim = ",", col_names = TRUE, col_types = NULL)

```

## Salaries by highest qualification
``` {r plotByHighestQual}

# Set factor levels
soccerSalariesData$highestQualCategory <- factor(
	soccerSalariesData$highestQualCategory,
	levels = c("PhD", "Postgraduate", "Undergraduate", "Unclear"))

# Create a list to store common theme options for use across plots
plotFeatures <- theme(
	plot.title = element_text(size = 16),
	plot.subtitle = element_text(size = 14),
	axis.title.x = element_blank(),
  axis.text.x = element_text(size = 11),
	axis.title.y = element_text(size = 16),
	axis.text.y = element_text(size = 11),
	legend.position = "none")

p <- ggplot(soccerSalariesData, aes(x = highestQualCategory, y = salaryGBP))
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = highestQualCategory))
p <- p + scale_y_continuous(
	limits = c(0, 50000),
	breaks = seq(0, 50000, by = 10000))
p <- p + labs(
	title = "UK soccer salaries of support staff",
	subtitle = "By highest qualification",
	y = "Salary (GBP)")
p <- p + plotFeatures
salaryByHighestQual <- p

# Display the plot
salaryByHighestQual

```

### Key findings

## What's next?

In the six and final post in this series, I will outline how a thoughtful data analysis and interpretation workflow provides hints as to the ways we can improve future data collection.

Constructive feedback and comments welcome - get in touch with me via [Twitter: @jacquietran](www.twitter.com/jacquietran)