---
title: How much do you get paid? Part IV - Academies by category status
author: Jacquie Tran
date: '2018-01-06'
slug: how-much-do-you-get-paid-part-4
categories:
  - Off-field analyses
tags:
  - football
  - salaries
  - soccer
  - sport science
  - strength and conditioning
draft: yes
---

## Introduction
This post is the fourth in a series - I suggest reading the **first**, **second**, and **third** posts for background. In this series, I will show how I created the plots that Aiden published on Twitter [here](www.twitter.com/ajoakley/status/944564698166513665) and [here](https://twitter.com/ajoakley/status/944936008285552642), alongside some commentary about the choices I made throughout the process and my own interpretations of the results.

## Getting started
Load the data and the libraries **as we've done before**.

``` {r globalOptions, echo = FALSE}

knitr::opts_chunk$set(fig.width = 7,
											fig.height = 5,
											fig.align = 'center',
											echo = TRUE,
											warning = FALSE,
											message = FALSE)

```

``` {r setup}

library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)

# Read the data into R
dataSource <- "https://raw.githubusercontent.com/jacquietran/soccerSalariesDataPublic/master/soccerSalariesDataPublic.csv"
soccerSalariesData <- read_delim(dataSource, delim = ",", col_names = TRUE, col_types = NULL)

```

## Salaries by academy category
The FA sets the academy status of all football clubs with academy programs. Budgets are somewhat dependent on guidelines that are category specific. How does academy category affect salary of strength and conditioning coaches and sport scientists?

``` {r plotByAcademyCategory}

# Subset of data frame to exclude academyCategory == NA
soccerSalariesData4 <- soccerSalariesData %>%
	filter(!is.na(academyCategory))

# Set factor levels
soccerSalariesData4$academyCategory <- factor(
	soccerSalariesData4$academyCategory,
	levels = c("1", "2", "3", "SFA Elite", "Development"))

# Create object to store axis labels with line breaks
labelsAcademyCategory <- c(
	"England\nCategory 1", "England\nCategory 2", "England\nCategory 3",
	"Scotland\nSFA Elite", "Scotland\nDevelopment")

# Create a list to store common theme options for use across plots
plotFeatures <- theme(
	plot.title = element_text(size = 16),
	plot.subtitle = element_text(size = 14),
	axis.title.x = element_blank(),
  axis.text.x = element_text(size = 11),
	axis.title.y = element_text(size = 16),
	axis.text.y = element_text(size = 11),
	legend.position = "none")

p <- ggplot(soccerSalariesData4, aes(x = academyCategory, y = salaryGBP))
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = academyCategory))
p <- p + scale_y_continuous(
	limits = c(0, 50000),
	breaks = seq(0, 50000, by = 10000))
p <- p + scale_x_discrete(labels = labelsAcademyCategory)
p <- p + labs(
	title = "UK soccer salaries of support staff: Academy staff only",
	subtitle = "By Academy Category",
	y = "Salary (GBP)")
p <- p + plotFeatures
salaryByAcademyCategory <- p

# Display the plot
salaryByAcademyCategory

```

### Key findings

## What's next?

In the fifth (and penultimate) post, I will investigate whether higher qualifications mean higher pay.

Constructive feedback and comments welcome - get in touch with me via [Twitter: @jacquietran](www.twitter.com/jacquietran)