---
title: How much do you get paid? Part III - Comparison by discipline
author: Jacquie Tran
date: '2017-01-05'
slug: how-much-do-you-get-paid-part-3
categories:
  - Random projects
tags:
  - sport science
  - strength and conditioning
  - football
  - salaries
  - soccer
draft: yes
---

## Introduction
This post is the third in a series - to get up to speed, here are some links so you can read **the first post** and **the second post**. In this series, I will show how I created the plots that Aiden published on Twitter [here](www.twitter.com/ajoakley/status/944564698166513665) and [here](https://twitter.com/ajoakley/status/944936008285552642), alongside some commentary about the choices I made throughout the process and my own interpretations of the results.

## Getting started
Load up the data and the libraries **as we've done before**.

``` {r globalOptions, echo = FALSE}

knitr::opts_chunk$set(fig.width = 7,
											fig.height = 5,
											fig.align = 'center',
											echo = TRUE,
											warning = FALSE,
											message = FALSE)

```

``` {r setup}

library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)

# Read the data into R
dataSource <- "https://raw.githubusercontent.com/jacquietran/soccerSalariesDataPublic/master/soccerSalariesDataPublic.csv"
soccerSalariesData <- read_delim(dataSource, delim = ",", col_names = TRUE, col_types = NULL)

```

## Salaries by league and by discipline
Let's compare how sport scientists and S&C coaches are paid in the different leagues.

``` {r plotByLeagueAndDiscipline}

# Subset of data frame to exclude dept == NA
soccerSalariesData3 <- soccerSalariesData %>%
	filter(!is.na(dept))
soccerSalariesData3$dept <- recode(
	soccerSalariesData3$dept,
	`Physical Performance` = "S&C")

# Set factor levels
soccerSalariesData3$dept <- factor(
	soccerSalariesData3$dept,
	levels = c("S&C", "Sport Science", "Both"))

# Create object to store axis labels with line breaks
labelsLeague <- c(
	"Premier\nLeague", "Scottish\nPremiership", "Championship", "League 1", "League 2")

# Create a list to store common theme options for use across plots
plotFeatures <- theme(
	plot.title = element_text(size = 16),
	plot.subtitle = element_text(size = 14),
	axis.title.x = element_blank(),
  axis.text.x = element_text(size = 11),
	axis.title.y = element_text(size = 16),
	axis.text.y = element_text(size = 11),
	legend.position = "none")

p <- ggplot(soccerSalariesData3, aes(x = leagueOfFirstTeam, y = salaryGBP))
p <- p + facet_wrap(~dept)
p <- p + geom_boxplot(outlier.shape = NA, coef = 0)
p <- p + geom_point(alpha = 1/4, size = 4, aes(colour = leagueOfFirstTeam))
p <- p + scale_y_continuous(
	limits = c(0, 60000),
	breaks = seq(0, 60000, by = 10000))
p <- p + scale_x_discrete(labels = labelsLeague)
p <- p + labs(
	title = "UK soccer salaries of support staff",
	subtitle = "By discipline",
	y = "Salary (GBP)")
p <- p + plotFeatures
salaryByLeagueAndDept <- p

# Display the plot
salaryByLeagueAndDept

```

### Key findings

## What's next?

In tomorrow's post, I'll take a look at salaries of football academy staff depending on their academy category.

Constructive feedback and comments welcome - get in touch with me via [Twitter: @jacquietran](www.twitter.com/jacquietran)