---
title: 'It''s #WBBL07 time! Analysing ball-by-ball data from past seasons'
author: Jacquie Tran
date: '2021-10-14'
slug: analysing-wbbl-ball-by-ball-data
categories:
  - On-field analyses
tags:
  - women's sport
  - r stats
  - sports analytics
  - cricket
draft: yes
---

# Getting started

I downloaded ball-by-ball match data from Cricsheet for the 6 preceding seasons of the Women's Big Bash League (264 matches). I chose to use the CSV "Ashwin" format.

```{r load_libraries, message = FALSE}

library(here)
library(readr)
library(dplyr)
library(stringr)
library(showtext)
library(ggplot2)
library(gghighlight)
library(ggtext)
library(paletteer)
library(patchwork)

```

```{r import_data}

all_matches <- read_csv(
  here("static/data/wbbl_01_to_06/all_matches.csv"),
  show_col_types = FALSE)

```

```{r add_google_fonts}

font_add_google("Roboto Condensed", "roboto_con")
font_add_google("Staatliches", "staat")

```

# Alyssa Healy

## Per season production

Alyssa Healy compared to all players who have batted in 3+ innings in a season.

Data from 2015/16 (WBBL01) excluded due to only having 10 matches worth of data in the Cricsheet spreadsheet.

```{r batting_per_season}

batting_per_season <- all_matches %>%
  # Exclude 2015/16 data - incomplete
  filter(season != "2015/16") %>%
  group_by(season, striker) %>%
  summarise(
    innings_total = length(unique(match_id)),
    runs_off_bat_total = sum(runs_off_bat),
    balls_faced_total = length(ball),
    dismissal_total = sum(!is.na(wicket_type)),
    .groups = "keep") %>%
  mutate(
    runs_per_innings_avg = round(runs_off_bat_total / innings_total, 1),
    batting_avg = round(runs_off_bat_total / dismissal_total, 1),
    strike_rate = round(runs_off_bat_total / balls_faced_total * 100, 1)) %>%
  filter(innings_total > 2)

```

```{r batting_per_season_healy}

batting_per_season_healy <- batting_per_season %>%
  mutate(
    is_healy = case_when(
      striker == "AJ Healy" ~ 1,
      TRUE                  ~ 0))

```

```{r plot_healy_per_season, warning = FALSE, fig.width = 10, fig.height = 8}

showtext_auto()

p <- ggplot(batting_per_season_healy,
            aes(x = season, y = runs_per_innings_avg, group = striker, colour = is_healy))
p <- p + geom_line(size = 3, colour = paletteer_d("beyonce::X25")[5])
p <- p + gghighlight(
  is_healy > 0,
  label_key = striker,
  label_params = aes(
    size = 16, force_pull = 0.1, nudge_y = 10, label.size = 2,
    family = "roboto_con", label.padding = 0.5,
    fill = paletteer_d("beyonce::X116")[4],
    colour = paletteer_d("beyonce::X25")[5]),
  unhighlighted_params = aes(size = 1),
  unhighlighted_colour = paletteer_d("beyonce::X73")[7])
p <- p + labs(
  title = "WBBL: Average runs scored per innings (3+ innings)",
  x = NULL, y = NULL,
  caption = "**Source:** Cricsheet.org // **Plot:** @jacquietran")
p <- p + theme_minimal()
p <- p + theme(
  text = element_text(
    size = 48, family = "roboto_con",
    colour = "#FFFFFF"),
  plot.title = element_text(family = "staat", margin=margin(0,0,15,0)),
  plot.caption = element_markdown(size = NULL, margin=margin(15,0,0,0)),
  axis.text = element_text(colour = "#FFFFFF"),
  legend.position = "none",
  panel.grid.major = element_line(linetype = "dashed"),
  panel.grid.minor = element_blank(),
  plot.background = element_rect(
      fill = paletteer_d("beyonce::X116")[3],
      colour = NA),
  panel.spacing = unit(2, "lines"),
  plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))
plot_healy_per_season <- p

plot_healy_per_season

ggsave(
  here("static/figs/runs_per_innings_healy.png"),
  last_plot(), device = "png", dpi = 300,
  width = 8, height = 6, units = "in")

showtext_auto(FALSE)

```

## Dismissals by ball number

```{r ball_number_faced}

# Create new variable for ball number in each over
ball_number_faced <- all_matches %>%
  filter(season != "2015/16") %>%
  mutate(
    ball_num_in_over = sub(".*\\.", "", ball))

# Summarise number of balls faced of each ball number, per batter
ball_number_faced_summary <- ball_number_faced %>%
  group_by(ball_num_in_over, striker) %>%
  summarise(balls_faced = n(), .groups = "keep") %>%
  ungroup()

# Dismissals by ball number
dismissals_by_ball_number <- ball_number_faced %>%
  select(ball_num_in_over, striker, wicket_type) %>%
  filter(!is.na(wicket_type)) %>%
  group_by(ball_num_in_over, striker) %>%
  summarise(dismissals_n = n(), .groups = "keep") %>%
  ungroup()

# Merge data
dismissals_by_ball_number_summary <- left_join(
  ball_number_faced_summary, dismissals_by_ball_number) %>%
  tidyr::replace_na(list(dismissals_n = 0)) %>%
  group_by(striker) %>%
  mutate(total_balls_faced = sum(balls_faced)) %>%
  ungroup() %>%
  mutate(dismissals_pct = round(dismissals_n / balls_faced * 100, 2)) %>%
  # Include those who have faced more than 200 balls total
  filter(total_balls_faced >= 200) %>%
  # Exclude balls beyond 6 - infrequent occurrences
  filter(ball_num_in_over < 7)

# Extract data for specific players
# Healy
dismissals_by_ball_number_summary_healy <- dismissals_by_ball_number_summary %>%
  filter(striker == "AJ Healy")

# Mooney
dismissals_by_ball_number_summary_mooney <- dismissals_by_ball_number_summary %>%
  filter(striker == "BL Mooney")

# Lanning
dismissals_by_ball_number_summary_lanning <- dismissals_by_ball_number_summary %>%
  filter(striker == "MM Lanning")

# Perry
dismissals_by_ball_number_summary_perry <- dismissals_by_ball_number_summary %>%
  filter(striker == "EA Perry")

# Devine
dismissals_by_ball_number_summary_devine <- dismissals_by_ball_number_summary %>%
  filter(striker == "SFM Devine")

# Knight
dismissals_by_ball_number_summary_knight <- dismissals_by_ball_number_summary %>%
  filter(striker == "HC Knight")

```

```{r plot_dismissals_by_ball_number}

showtext_auto()

plot_features <- list(
  coord_cartesian(ylim = c(0, 10)),
  theme_minimal(),
  theme(
    text = element_text(
      size = 48, family = "roboto_con",
      colour = "#FFFFFF"),
    plot.title = element_text(family = "staat", margin=margin(0,0,15,0)),
    plot.subtitle = element_text(family = "staat", margin=margin(0,0,15,0)),
    plot.caption = element_markdown(size = NULL, margin=margin(15,0,0,0)),
    axis.text = element_text(colour = "#FFFFFF"),
    legend.position = "none",
    panel.grid.major.y = element_line(linetype = "dashed"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(
      fill = paletteer_d("beyonce::X116")[3],
      colour = NA),
    panel.spacing = unit(2, "lines"),
    plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))
  )

# Healy
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 2, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_healy,
  colour = paletteer_d("beyonce::X25")[5], size = 5.5)
p <- p + labs(
  title = "WBBL: % dismissals by ball number",
  subtitle = "AJ Healy",
  x = NULL, y = NULL)
p <- p + plot_features
p1 <- p
# plot_dismissals_by_ball_num_healy <- p

# Mooney
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 2, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_mooney,
  colour = paletteer_d("NineteenEightyR::miami1")[5], size = 5.5)
p <- p + labs(
  subtitle = "BL Mooney",
  x = NULL, y = NULL)
p <- p + plot_features
p2 <- p

# Lanning
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 2, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_lanning,
  colour = paletteer_d("beyonce::X93")[31], size = 5.5)
p <- p + labs(
  subtitle = "MM Lanning",
  x = NULL, y = NULL)
p <- p + plot_features
p3 <- p

# Perry
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 2, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_perry,
  colour = paletteer_d("beyonce::X25")[5], size = 5.5)
p <- p + labs(
  subtitle = "EA Perry",
  # caption = "",
  x = NULL, y = NULL)
p <- p + plot_features
p4 <- p

# Devine
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 2, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_devine,
  colour = paletteer_d("NineteenEightyR::miami1")[5], size = 5.5)
p <- p + labs(
  subtitle = "SFM Devine",
  # caption = "",
  x = NULL, y = NULL)
p <- p + plot_features
p5 <- p

# Knight
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 2, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_knight,
  colour = paletteer_d("LaCroixColoR::Lime")[2], size = 5.5)
p <- p + labs(
  subtitle = "HC Knight",
  caption = "**Source:** Cricsheet.org // **Plot:** @jacquietran",
  x = NULL, y = NULL)
p <- p + plot_features
p6 <- p

showtext_auto(FALSE)

```

```{r quilt_dismissals_by_ball_number}

quilt_dismissals_by_ball_number <- (p1 + p2 + p3) / (p4 + p5 + p6)

ggsave(
  here("static/figs/dismissals_by_ball_num.png"),
  last_plot(), device = "png", dpi = 300,
  width = 12, height = 8, units = "in")

```
