---
title: 'EXPLORATION: It''s #WBBL07 time! Analysing ball-by-ball data from past seasons'
author: Jacquie Tran
date: '2021-10-14'
slug: analysing-wbbl-ball-by-ball-data
categories:
  - On-field analyses
tags:
  - women's sport
  - r stats
  - sports analytics
  - cricket
draft: yes
---

# Getting started

I downloaded ball-by-ball match data from Cricsheet for the 6 preceding seasons of the Women's Big Bash League (264 matches). I chose to use the CSV "Ashwin" format.

```{r load_libraries, message = FALSE}

library(here)
library(readr)
library(dplyr)
library(ggplot2)

```

```{r import_data}

all_matches <- read_csv(
  here("static/data/wbbl_01_to_06/all_matches.csv"),
  show_col_types = FALSE)

```

# Batting

## Alyssa Healy

```{r batting_healy}

batting_healy <- all_matches %>%
  filter(striker == "AJ Healy")

```

Batting totals

```{r batting_totals_healy}

batting_totals_healy <- batting_healy %>%
  summarise(
    innings_total = length(unique(match_id)),
    runs_off_bat_total = sum(runs_off_bat),
    dismissal_total = sum(!is.na(wicket_type)))

```

Per season totals

```{r batting_per_season_healy}

batting_per_season_healy <- batting_healy %>%
  group_by(season) %>%
  summarise(
    innings_total = length(unique(match_id)),
    runs_off_bat_total = sum(runs_off_bat),
    dismissal_total = sum(!is.na(wicket_type))) %>%
  mutate(
    runs_per_innings_avg = runs_off_bat_total / innings_total,
    batting_avg = runs_off_bat_total / dismissal_total)

```

Most frequent modes of dismissal

```{r batting_healy_dismissals}

dismissal_summary_healy <- batting_healy %>%
  filter(!is.na(wicket_type)) %>%
  group_by(wicket_type) %>%
  summarise(n = n())

```

Wickets lost to specific bowlers

```{r}

wickets_to_bowlers_healy <- batting_healy %>%
  filter(!is.na(wicket_type)) %>%
  group_by(bowler) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

```

## Sophie Devine

```{r batting_devine}

batting_devine <- all_matches %>%
  filter(striker == "SFM Devine")

```

Batting totals

```{r batting_totals_devine}

batting_totals_devine <- batting_devine %>%
  summarise(
    innings_total = length(unique(match_id)),
    runs_off_bat_total = sum(runs_off_bat),
    dismissal_total = sum(!is.na(wicket_type)))

```

```{r batting_per_season_devine}

batting_per_season_devine <- batting_devine %>%
  group_by(season) %>%
  summarise(
    innings_total = length(unique(match_id)),
    runs_off_bat_total = sum(runs_off_bat),
    dismissal_total = sum(!is.na(wicket_type))) %>%
  mutate(
    runs_per_innings_avg = runs_off_bat_total / innings_total,
    batting_avg = runs_off_bat_total / dismissal_total)

```

# Bowling

## Molly Strano

## Annabel Sutherland