---
title: '#WBBL visualisations'
author: Jacquie Tran
date: '2021-11-10'
slug: wbbl-visualisations
categories:
  - On-field analyses
tags:
  - women's sport
  - r stats
  - sports analytics
  - cricket
draft: yes
---

# Introduction

It's been a thrilling #WBBL07 so far - the competition seems well-balanced, with every team boasting world-class athletes alongside the young guns who are the next wave of Australian talent coming through.

Coming off the back of the hotly contested Australia vs. India series, I have been glued to the TV. I thought I might as well turn my attention to another screen instead - my computer screen - and take a closer look at some key performances so far this season.

Here's a little tour of some visualisations I've made recently using Women's Big Bash League (WBBL) data.

# Getting started

I use a few custom functions to retrieve data from [Cricsheet.org](https://cricsheet.org), and I'm continuing to work on them over [here](https://github.com/jacquietran/cricsheet). *(Buyer beware: they work if you know the right sequence of actions and say some nice things in the general direction of your RStudio IDE.)*

```{r load_libraries, message = FALSE}

library(here)
library(readr)
library(dplyr)
library(stringr)
library(showtext)
library(ggplot2)
library(gghighlight)
library(ggtext)
library(paletteer)
library(patchwork)

```

```{r import_csv, echo = FALSE, warning = FALSE}

# Import Cricsheet data from static CSVs
wbbl_bbb <- readr::read_csv(
  here::here("static/data/wbbl/wbbl_bbb.csv"),
  show_col_types = FALSE)

wbbl_match_info <- readr::read_csv(
  here::here("static/data/wbbl/wbbl_match_info.csv"),
  show_col_types = FALSE)

```


```{r retrieve_cricsheet_data, eval = FALSE}

# Fetch ball-by-ball data
wbbl_bbb <- fetch_cricsheet_bbb("wbbl")

# Fetch match metadata
wbbl_match_info <- fetch_cricsheet_match_info("wbbl")

```

## Alyssa Healy's WBBL batting record

What is there to say about Alyssa Healy that hasn't already been written by a far better writer than I? (If you haven't seen it, check out [this recent feature article by Adam Burnett](https://www.cricket.com.au/news/feature/alyssa-healy-australia-india-mitch-starc-vice-captain-sister-kareen-family-lanning-mott/2021-10-10).)

She's a superstar at every level she plays at, so I wanted to visualise her production across seasons, compared to other batters in the WBBL.

### Tidying the data

On exploring the Cricsheet data, I noticed that there are only 10 matches from WBBL01 with ball-by-ball data, but there were 59 matches played in the 1st season of WBBL (per [Wikipedia](https://en.wikipedia.org/wiki/2015%E2%80%9316_Women%27s_Big_Bash_League_season)).

Additionally, when I originally worked on this plot of [Healy vs. the world](https://twitter.com/jacquietran/status/1448534315202080773), it was mid-October and the first ball of #WBBL07 had yet to be bowled. Now, we are reproducing the chart as we are part-way through the current season, so the code below also omits data from any matches played so far in this 7th season.

```{r tidy_bbb_data}

# Data from 2015/16 (WBBL01) excluded due to only having 10 matches worth of data in the Cricsheet spreadsheet.
# Data from 2021/22 (WBBL07) excluded as season is currently in progress.
wbbl_bbb %>%
  filter(!season %in% c("2015/16", "2021/22")) -> wbbl_bbb_tidy

```

The ball-by-ball data hosted by Cricsheet provides a great starting point, but a little more tidying and wrangling is needed for the purposes of understanding batting performances across WBBL seasons.

In cricket broadcasts, batting average is probably the most common statistic that commentators will reference to highlight how well a player performs with willow in hand. Batting average is calculated by taking runs scored over a period of time and dividing it by the number of dismissals (e.g., )

Given the T20 format of the WBBL, I was more interested in 

```{r batting_per_season_healy}

# Alyssa Healy compared to all players who have batted in 3+ innings in a season.
wbbl_bbb_tidy %>%
  group_by(season, striker) %>%
  summarise(
    innings_total = length(unique(match_id)),
    runs_off_bat_total = sum(runs_off_bat),
    balls_faced_total = length(ball),
    .groups = "keep") %>%
  mutate(
    runs_per_innings_avg = round(runs_off_bat_total / innings_total, 1),
    strike_rate = round(runs_off_bat_total / balls_faced_total * 100, 1)) %>%
  filter(innings_total > 2) -> batting_per_season

# Flag Healy's per season records
batting_per_season %>%
  mutate(
    is_healy = case_when(
      striker == "AJ Healy" ~ 1,
      TRUE                  ~ 0)) -> batting_per_season_healy

```

```{r add_google_fonts}

font_add_google("Roboto Condensed", "roboto_con")
font_add_google("Staatliches", "staat")

```

```{r plot_healy_per_season, warning = FALSE, fig.height = 6, fig.width = 8}

showtext_auto()

p <- ggplot(batting_per_season_healy,
            aes(x = season, y = runs_per_innings_avg,
                group = striker, colour = is_healy))
p <- p + geom_line(size = 3, colour = paletteer_d("beyonce::X25")[5])
p <- p + gghighlight(
  is_healy > 0,
  label_key = striker,
  label_params = aes(
    size = 6, force_pull = 0.1, nudge_y = 10, label.size = 2,
    family = "roboto_con", label.padding = 0.5,
    fill = paletteer_d("beyonce::X116")[4],
    colour = paletteer_d("beyonce::X25")[5]),
  unhighlighted_params = aes(size = 1),
  unhighlighted_colour = paletteer_d("beyonce::X73")[7])
p <- p + labs(
  title = "WBBL: Average runs scored per innings (3+ innings)",
  x = NULL, y = NULL,
  caption = "**Source:** Cricsheet.org // **Plot:** @jacquietran")
p <- p + theme_minimal()
p <- p + theme(
  text = element_text(
    size = 18, family = "roboto_con", colour = "#FFFFFF"),
  plot.title = element_text(family = "staat", margin=margin(0,0,15,0)),
  plot.caption = element_markdown(size = NULL, margin=margin(15,0,0,0)),
  axis.text = element_text(colour = "#FFFFFF"),
  legend.position = "none",
  panel.grid.major = element_line(linetype = "dashed"),
  panel.grid.minor = element_blank(),
  plot.background = element_rect(
      fill = paletteer_d("beyonce::X116")[3],
      colour = NA),
  panel.spacing = unit(2, "lines"),
  plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))
plot_healy_per_season <- p

plot_healy_per_season

```

```{r echo = FALSE}

showtext_auto(FALSE)

```

## Dismissals by ball number

```{r ball_number_faced}

# Create new variable for ball number in each over
ball_number_faced <- wbbl_bbb_tidy %>%
  mutate(
    ball_num_in_over = sub(".*\\.", "", ball))

# Summarise number of balls faced of each ball number, per batter
ball_number_faced_summary <- ball_number_faced %>%
  group_by(ball_num_in_over, striker) %>%
  summarise(balls_faced = n(), .groups = "keep") %>%
  ungroup()

# Dismissals by ball number
dismissals_by_ball_number <- ball_number_faced %>%
  select(ball_num_in_over, striker, wicket_type) %>%
  filter(!is.na(wicket_type)) %>%
  group_by(ball_num_in_over, striker) %>%
  summarise(dismissals_n = n(), .groups = "keep") %>%
  ungroup()

# Merge data
dismissals_by_ball_number_summary <- left_join(
  ball_number_faced_summary, dismissals_by_ball_number) %>%
  tidyr::replace_na(list(dismissals_n = 0)) %>%
  group_by(striker) %>%
  mutate(total_balls_faced = sum(balls_faced)) %>%
  ungroup() %>%
  mutate(dismissals_pct = round(dismissals_n / balls_faced * 100, 2)) %>%
  # Include those who have faced more than 200 balls total
  filter(total_balls_faced >= 200) %>%
  # Exclude balls beyond 6 - infrequent occurrences
  filter(ball_num_in_over < 7)

# Extract data for specific players
# Healy
dismissals_by_ball_number_summary_healy <- dismissals_by_ball_number_summary %>%
  filter(striker == "AJ Healy")

# Mooney
dismissals_by_ball_number_summary_mooney <- dismissals_by_ball_number_summary %>%
  filter(striker == "BL Mooney")

# Lanning
dismissals_by_ball_number_summary_lanning <- dismissals_by_ball_number_summary %>%
  filter(striker == "MM Lanning")

# Perry
dismissals_by_ball_number_summary_perry <- dismissals_by_ball_number_summary %>%
  filter(striker == "EA Perry")

# Devine
dismissals_by_ball_number_summary_devine <- dismissals_by_ball_number_summary %>%
  filter(striker == "SFM Devine")

# Knight
dismissals_by_ball_number_summary_knight <- dismissals_by_ball_number_summary %>%
  filter(striker == "HC Knight")

```

```{r plot_dismissals_by_ball_number}

plot_features <- list(
  coord_cartesian(ylim = c(0, 10)),
  theme_minimal(),
  theme(
    text = element_text(
      size = 18, family = "roboto_con",
      colour = "#FFFFFF"),
    plot.title = element_text(family = "staat", margin=margin(0,0,15,0)),
    plot.subtitle = element_text(family = "staat", margin=margin(0,0,15,0)),
    plot.caption = element_markdown(size = NULL, margin=margin(15,0,0,0)),
    axis.text = element_text(colour = "#FFFFFF"),
    legend.position = "none",
    panel.grid.major.y = element_line(linetype = "dashed"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(
      fill = paletteer_d("beyonce::X116")[3],
      colour = NA),
    panel.spacing = unit(2, "lines"),
    plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))
  )

showtext_auto()

# Healy
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 1, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_healy,
  colour = paletteer_d("beyonce::X25")[5], size = 3)
p <- p + labs(
  title = "WBBL: % dismissals by ball number",
  subtitle = "AJ Healy",
  x = NULL, y = NULL)
p <- p + plot_features
p1 <- p
# plot_dismissals_by_ball_num_healy <- p

# Mooney
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 1, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_mooney,
  colour = paletteer_d("NineteenEightyR::miami1")[5], size = 3)
p <- p + labs(
  subtitle = "BL Mooney",
  x = NULL, y = NULL)
p <- p + plot_features
p2 <- p

# Lanning
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 1, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_lanning,
  colour = paletteer_d("beyonce::X93")[31], size = 3)
p <- p + labs(
  subtitle = "MM Lanning",
  x = NULL, y = NULL)
p <- p + plot_features
p3 <- p

# Perry
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 1, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_perry,
  colour = paletteer_d("beyonce::X25")[5], size = 3)
p <- p + labs(
  subtitle = "EA Perry",
  # caption = "",
  x = NULL, y = NULL)
p <- p + plot_features
p4 <- p

# Devine
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 1, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_devine,
  colour = paletteer_d("NineteenEightyR::miami1")[5], size = 3)
p <- p + labs(
  subtitle = "SFM Devine",
  # caption = "",
  x = NULL, y = NULL)
p <- p + plot_features
p5 <- p

# Knight
p <- ggplot(dismissals_by_ball_number_summary,
            aes(x = ball_num_in_over, y = dismissals_pct))
p <- p + geom_boxplot(
  fill = "#FFFFFF", colour = "#FFFFFF", size = 1, alpha = 0.25,
  notch = TRUE, outlier.shape = NA, coef = 0)
p <- p + geom_point(
  data = dismissals_by_ball_number_summary_knight,
  colour = paletteer_d("LaCroixColoR::Lime")[2], size = 3)
p <- p + labs(
  subtitle = "HC Knight",
  caption = "**Source:** Cricsheet.org // **Plot:** @jacquietran",
  x = NULL, y = NULL)
p <- p + plot_features
p6 <- p

```

```{r quilt_dismissals_by_ball_number}

# Quilt the plots
quilt_dismissals_by_ball_number <- (p1 + p2 + p3) / (p4 + p5 + p6)

# Display the quilted plot
quilt_dismissals_by_ball_number

```

```{r}

showtext_auto(FALSE)

```

## Wiggly bois

```{r}

wbbl_match_info %>%
  filter(season == "2021/22") %>%
  select(
    match_id, winner, winner_runs, winner_wickets, method, outcome,
    eliminator) %>%
  mutate(
    match_id = factor(match_id)) -> wbbl07_match_info_tidy

wbbl_bbb %>%
  filter(season == "2021/22") %>%
  tidyr::separate(ball, c("over_num_extracted", "ball_num")) %>%
  mutate(
    match_id = factor(match_id),
    over_num_numeric = as.numeric(over_num_extracted) + 1,
    runs_scored = runs_off_bat + extras,
    ball_num = as.numeric(ball_num),
    wicket_type = case_when(
      wicket_type == "" ~ NA_character_,
      TRUE              ~ wicket_type)) %>%
  group_by(match_id, innings) %>%
  mutate(
    temp_var = 1,
    balls_cumulative = cumsum(temp_var),
    runs_cumulative = cumsum(runs_scored),
    runs_total = max(runs_cumulative)) %>%
  ungroup() %>%
  select(-temp_var) %>%
  left_join(., wbbl07_match_info_tidy) %>%
  mutate(
    outcome_batting_team = case_when(
      outcome %in% c("no result", "tie") ~ as.character(outcome),
      winner == batting_team             ~ "won",
      TRUE                               ~ "lost"),
    outcome_bowling_team = case_when(
      outcome %in% c("no result", "tie") ~ as.character(outcome),
      winner == bowling_team             ~ "won",
      TRUE                               ~ "lost")
    ) -> wbbl07_bbb_tidy

```

```{r team_strike_rate}

wbbl07_bbb_tidy %>%
  # Exclude matches that ended with a Super Over ("tie")
  # and matches that were called off ("no result")
  filter(!outcome_batting_team %in% c("tie", "no result")) %>%
  group_by(match_id, innings) %>%
  mutate(
    rolling_strike_rate = round(
      runs_cumulative / balls_cumulative * 100, 1),
    wicket_ball_num = case_when(
      !is.na(wicket_type) ~ balls_cumulative,
      TRUE                ~ NA_real_),
    wicket_strike_rate = case_when(
      !is.na(wicket_type) ~ rolling_strike_rate,
      TRUE                ~ NA_real_),
    innings_description = case_when(
      innings == 1 ~ "Batting 1st",
      innings == 2 ~ "Batting 2nd"),
    bowling_team_short = word(bowling_team, -1),
    start_date_day = lubridate::day(start_date),
    start_date_month = lubridate::month(start_date),
    match_details = glue::glue(
      "{innings_description} vs. {bowling_team_short} ({start_date_day}/{start_date_month})")) %>%
  arrange(match_id, innings, balls_cumulative) %>%
  mutate(
    match_details = factor(
      match_details, levels = unique(match_details)),
    outcome_batting_team = factor(
      outcome_batting_team, levels = c("won", "lost")
    )) -> team_strike_rate

```


```{r team_strike_rate_renegades, fig.height = 10, fig.width = 8, warning = FALSE}

# Filter to Renegades' innings only
team_strike_rate %>%
  filter(
    str_detect(batting_team, "Renegades")
    ) -> team_strike_rate_renegades

showtext_auto()

# Build plot
p <- ggplot(
  team_strike_rate_renegades,
  aes(x = balls_cumulative, y = rolling_strike_rate))
p <- p + facet_wrap(~ match_details, ncol = 3)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team))
p <- p + geom_point(
  aes(x = team_strike_rate_renegades$wicket_ball_num,
      y = team_strike_rate_renegades$wicket_strike_rate),
      colour = "red", size = 1, alpha = 0.75)
p <- p + labs(
  title = "WBBL07: Melbourne Renegades - Team strike rate (games up to 7 Nov 2021)",
  x = "Ball number in an innings", y = NULL,
  caption = "**Source:** Cricsheet.org // **Plot:** @jacquietran")
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + scale_color_manual(
  values = c("won" = "#4a8bad", "lost" = "#AD4A8B"),
  labels = c("Renegades won", "Renegades lost"))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  text = element_text(
    size = 18, family = "roboto_con",
    colour = "#FFFFFF"),
  legend.position = "top",
  legend.title = element_blank(),
  legend.key.size = unit(1.5, "cm"),
  legend.margin = margin(0, 0, 0, 0),
  legend.spacing.x = unit(0, "cm"),
  legend.spacing.y = unit(0, "cm"),
  plot.title = element_text(family = "staat", margin=margin(0,0,15,0)),
  plot.caption = element_markdown(size = NULL, margin=margin(15,0,0,0)),
  strip.text = element_text(colour = "#FFFFFF", size = 12),
  axis.text = element_text(colour = "#FFFFFF"),
  axis.title.x = element_text(margin=margin(15,0,0,0)),
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.major.x = element_line(
    colour = "#203b60", linetype = "dotted"),
  panel.grid.minor.x = element_blank(),
  plot.background = element_rect(
      fill = paletteer_d("beyonce::X116")[3],
      colour = NA),
  panel.spacing = unit(2, "lines"),
  plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))
plot_team_strike_rate_renegades <- p

# Display plot
plot_team_strike_rate_renegades

```

```{r}

showtext_auto(FALSE)

```