---
title: '#WBBL07: Racing towards finals'
author: Jacquie Tran
date: '2021-11-09'
slug: wbbl07-racing-towards-finals
categories:
  - On-field analyses
tags:
  - women's sport
  - r stats
  - sports analytics
  - cricket
draft: yes
---

# Introduction

It's been a thrilling #WBBL07 so far - the competition seems well-balanced, with every team boasting world-class athletes alongside the young guns who are the next wave of Australian talent coming through.

Coming off the back of the hotly contested Australia vs. India series, I have been glued to the TV and thought I might as well turn my attention to another screen instead - my computer screen - and take a closer look at some key performances so far this season.

# Getting started

I downloaded ball-by-ball data and match metadata from Cricsheet for the Women's Big Bash League (specifically using the CSV "Ashwin" format), focusing on games played so far in season 7 of the Women's Big Bash League (32 matches played between Oct 14 and Nov 7 2021, excluding the match called off between the Hobart Hurricanes and the Adelaide Strikers).

```{r load_libraries, message = FALSE}

library(here)
library(readr)
library(dplyr)
library(stringr)
library(showtext)
library(ggplot2)
library(gghighlight)
library(ggtext)
library(paletteer)
library(patchwork)

```

```{r import_data, message = FALSE}

# Run custom functions to fetch data from Cricsheet.org ------------------------

wbbl07_bbb <- fetch_cricsheet_bbb("wbbl") %>%
  filter(season == "2021/22")
wbbl07_match_info <- fetch_cricsheet_match_info("wbbl") %>%
  filter(season == "2021/22")

# Do a little tidying up first -------------------------------------------------

wbbl07_match_info %>%
  select(
    match_id, winner, winner_runs, winner_wickets, method, outcome,
    eliminator) %>%
  mutate(
    match_id = factor(match_id)) -> wbbl07_match_info_tidy

wbbl07_bbb %>%
  tidyr::separate(ball, c("over_num_extracted", "ball_num")) %>%
  mutate(
    match_id = factor(match_id),
    over_num_numeric = as.numeric(over_num_extracted) + 1,
    runs_scored = runs_off_bat + extras,
    ball_num = as.numeric(ball_num),
    wicket_type = case_when(
      wicket_type == "" ~ NA_character_,
      TRUE              ~ wicket_type)) %>%
  group_by(match_id, innings) %>%
  mutate(
    temp_var = 1,
    balls_cumulative = cumsum(temp_var),
    runs_cumulative = cumsum(runs_scored),
    runs_total = max(runs_cumulative)) %>%
  ungroup() %>%
  select(-temp_var) %>%
  left_join(., wbbl07_match_info_tidy) %>%
  mutate(
    outcome_batting_team = case_when(
      outcome %in% c("no result", "tie") ~ as.character(outcome),
      winner == batting_team             ~ "won",
      TRUE                               ~ "lost"),
    outcome_bowling_team = case_when(
      outcome %in% c("no result", "tie") ~ as.character(outcome),
      winner == bowling_team             ~ "won",
      TRUE                               ~ "lost")
    ) -> wbbl07_bbb_tidy

```

# Team strike rate after wickets

One tactical consideration that makes the shorter formats so fascinating is how to go about scoring runs after you've lost a wicket.

With only 120 regulation balls bowled in a 20-over match:

- How many balls do you take to get your eye in?
- When do you decide to push on and swing the bat?
- How does this differ with "game state", when the wickets and overs remaining equation changes?

One way we can look at this is to examine patterns in team strike rate (runs scored for balls faced) across the course of an innings.

```{r team_strike_rate}

wbbl07_bbb_tidy %>%
  # Exclude matches that ended with a Super Over ("tie")
  # and matches that were called off ("no result")
  filter(!outcome_batting_team %in% c("tie", "no result")) %>%
  group_by(match_id, innings) %>%
  mutate(
    rolling_strike_rate = round(
      runs_cumulative / balls_cumulative * 100, 1),
    wicket_ball_num = case_when(
      !is.na(wicket_type) ~ balls_cumulative,
      TRUE                ~ NA_real_),
    wicket_strike_rate = case_when(
      !is.na(wicket_type) ~ rolling_strike_rate,
      TRUE                ~ NA_real_),
    innings_description = case_when(
      innings == 1 ~ "Batting 1st",
      innings == 2 ~ "Batting 2nd"),
    bowling_team_short = word(bowling_team, -1),
    start_date_day = lubridate::day(start_date),
    start_date_month = lubridate::month(start_date),
    match_details = glue::glue(
      "{innings_description} vs. {bowling_team_short} ({start_date_day}/{start_date_month})")) %>%
  arrange(match_id, innings, balls_cumulative) %>%
  mutate(
    match_details = factor(
      match_details, levels = unique(match_details)),
    outcome_batting_team = factor(
      outcome_batting_team, levels = c("won", "lost")
    )) -> team_strike_rate

```

## Plot: Team strike rate from winning vs. losing teams

Firstly, I looked at team strike rate when the batting team 1) won or lost, and 2) batted first (set a target) or batted second (chased a target).

```{r add_google_fonts}

font_add_google("Roboto Condensed", "roboto_con")
font_add_google("Staatliches", "staat")

```

```{r plot_team_strike_rate_by_outcome, message = FALSE}

# Build plot
p <- ggplot(
  team_strike_rate,
  aes(x = balls_cumulative, y = rolling_strike_rate, group = match_id))
p <- p + facet_grid(innings_description ~ outcome_batting_team)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team))
p <- p + geom_smooth(aes(group = outcome_batting_team), method = "loess")
#p <- p + geom_point(
#  aes(x = team_strike_rate$wicket_ball_num,
#      y = team_strike_rate$wicket_strike_rate),
#      colour = "red", size = 3, alpha = 0.75)
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  legend.position = "none",
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank())
plot_team_strike_rate_by_outcome <- p

# Display plot
plot_team_strike_rate_by_outcome

```
First things first - the sample size is not huge. The plot covers 60 innings total from 30 matches played to this point in the season (excluding tied matches that went to Super Overs and the match called off due to rain), so there are 15 innings in each panel.

Charting team strike rate for each innings in addition to the local regression trendlines illustrates the degree of variation in strike rate "patterns", so we don't want to be too conclusive with our judgments based on the LOESS curve alone.

Some reckons and questions:

- For teams that won after batting first:
     - They have typically started off slowly (given the T20 format of WBBL), only achieving parity in team strike rate at somewhere round the 5-over mark.
     - Once they get to scoring about a run a ball (across the innings), they keep accelerating through overs 5 to 10, and then hold their team strike rate steady at ~115-120 (i.e., ~1.15 to 1.20 runs per ball) until the close of innings.
- For teams that have lost (irrespective of batting first or second), they only manage to get to strike rate parity very late in the innings, somewhere around the 18th or 19th over.
- As to be expected, the LOESS curve for winning teams that batted first looks similar to the curve for losing teams that batted second. The target set by the first team to bat dictates how the chasing team will bat too.

## Plot: Team strike rate by team

```{r plot_team_strike_rate_by_team, message = FALSE}

# Build plot
p <- ggplot(
  team_strike_rate,
  aes(x = balls_cumulative, y = rolling_strike_rate, group = match_id))
p <- p + facet_grid(outcome_batting_team ~ batting_team)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team))
p <- p + geom_smooth(aes(group = outcome_batting_team), method = "loess")
#p <- p + geom_point(
#  aes(x = team_strike_rate$wicket_ball_num,
#      y = team_strike_rate$wicket_strike_rate),
#      colour = "red", size = 3, alpha = 0.75)
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  legend.position = "none",
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank())
plot_team_strike_rate_by_team <- p

# Display plot
plot_team_strike_rate_by_team

```

The small sample caveat is even more important here - for instance, Sydney Thunder have only won two matches, so the local regression curve is based off only two innings batted in matches they've won.

Some possible patterns:

- When the Scorchers get set, they can really unleash.
- The Sixers have been a two-speed team so far this season; when it has come together for them, they have looked the part and fulfilled the promise of their batting line up. What surprised me was to see that their strike rate pattern in losing matches looks quite similar to bottom-of-the-ladder teams Hobart Hurricanes and Sydney Thunder.
- The current league leaders, the Melbourne Renegades, don't show the same steep strike rate curves as teams like Brisbane and Perth that have noted power hitters.
- In wins and losses, the Hobart Hurricanes have struggled to generate the big scores that other teams have achieved. But there are 3 wins in there, which suggests that it is the strength of their bowling attack (featuring Molly Strano, Tayla Vlaeminck, and Nicola Carey) that is powering their victories.
- It's easy for the eye to be drawn to the prowess of Devine / Mooney, but the Harris / Redmayne combination over at Brisbane has been similarly destructive. As a team, Brisbane does not spend long striking at less than a run a ball.

## Team strike rate with wickets included

Plotting batting data is interesting and all, but cricket is a game between bat and ball - so to get a clearer picture of batting performance requires incorporating information about the bowling team's performance.

### Melbourne Renegades

Let's take a look at the Melbourne Renegades, who currently sit at the top of the standings (as of Nov 7 2021):

```{r plot_team_strike_rate_renegades}

# Filter to Renegades' innings only
team_strike_rate %>%
  filter(
    str_detect(batting_team, "Renegades")
    ) -> team_strike_rate_renegades

# Build plot
showtext_auto()

p <- ggplot(
  team_strike_rate_renegades,
  aes(x = balls_cumulative, y = rolling_strike_rate))
p <- p + facet_wrap(~ match_details, ncol = 3)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team), size = 1.5)
p <- p + geom_point(
  aes(x = team_strike_rate_renegades$wicket_ball_num,
      y = team_strike_rate_renegades$wicket_strike_rate),
      colour = "red", size = 3, alpha = 0.75)
p <- p + labs(
  title = "WBBL07: Melbourne Renegades - Team strike rate (games up to 7 Nov 2021)",
  x = "Ball number in an innings", y = NULL,
  caption = "**Source:** Cricsheet.org // **Plot:** @jacquietran")
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + scale_color_manual(
  values = c("won" = "#4a8bad", "lost" = "#AD4A8B"),
  labels = c("Renegades won", "Renegades lost"))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  text = element_text(
    size = 48, family = "roboto_con",
    colour = "#FFFFFF"),
  legend.position = "top",
  legend.title = element_blank(),
  legend.key.size = unit(1.5, "cm"),
  legend.margin = margin(0, 0, 0, 0),
  legend.spacing.x = unit(0, "cm"),
  legend.spacing.y = unit(0, "cm"),
  plot.title = element_text(family = "staat", margin=margin(0,0,15,0)),
  plot.caption = element_markdown(size = NULL, margin=margin(15,0,0,0)),
  strip.text = element_text(colour = "#FFFFFF", size = 32),
  axis.text = element_text(colour = "#FFFFFF"),
  axis.title.x = element_text(margin=margin(15,0,0,0)),
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.major.x = element_line(
    colour = "#203b60", linetype = "dotted"),
  panel.grid.minor.x = element_blank(),
  plot.background = element_rect(
      fill = paletteer_d("beyonce::X116")[3],
      colour = NA),
  panel.spacing = unit(2, "lines"),
  plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))
plot_team_strike_rate_renegades <- p

# Display plot
plot_team_strike_rate_renegades

# Export to file
ggsave(
  here("static/figs/team_strike_rate_renegades.png"),
  last_plot(), device = "png", dpi = 300,
  width = 8, height = 8, units = "in")

showtext_auto(FALSE)

```

### Sydney Thunder

And now let's plot the same variables but for the Sydney Thunder, who have had a few losses to start their title defense and are currently at the bottom of the standings on net run rate:

```{r plot_team_strike_rate_thunder}

# Filter to Thunder innings only
team_strike_rate %>%
  filter(
    str_detect(batting_team, "Thunder")
    ) -> team_strike_rate_thunder

# Build plot
p <- ggplot(
  team_strike_rate_thunder,
  aes(x = balls_cumulative, y = rolling_strike_rate))
p <- p + facet_wrap(~ match_details, ncol = 4)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team))
p <- p + geom_point(
  aes(x = team_strike_rate_thunder$wicket_ball_num,
      y = team_strike_rate_thunder$wicket_strike_rate),
      colour = "red", size = 3, alpha = 0.75)
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank())
plot_team_strike_rate_thunder <- p

# Display plot
plot_team_strike_rate_thunder

```