---
title: '#WBBL07: Racing towards finals'
author: Jacquie Tran
date: '2021-11-06'
slug: wbbl07-racing-towards-finals
categories:
  - On-field analyses
tags:
  - women's sport
  - r stats
  - sports analytics
  - cricket
draft: yes
---

# Introduction

It's been a thrilling #WBBL07 so far - the competition seems well-balanced, with every team boasting world-class athletes alongside the young guns who are the next wave of Australian talent coming through.

Coming off the back of the hotly contested Australia vs. India series, I have been glued to the TV and thought I might as well turn my attention to another screen instead - my computer screen - and take a closer look at some key performances so far this season.

# Getting started

I downloaded ball-by-ball data and match metadata from Cricsheet for the Women's Big Bash League (specifically using the CSV "Ashwin" format), focusing on games played so far in season 7 of the Women's Big Bash League (27 matches played between Oct 14 and Nov 3 2021).

```{r load_libraries, message = FALSE}

knitr::opts_chunk$set(eval = FALSE)

library(here)
library(readr)
library(dplyr)
library(stringr)
library(showtext)
library(ggplot2)
library(gghighlight)
library(ggtext)
library(paletteer)
library(patchwork)

```

```{r import_data, message = FALSE}

# Run custom functions to fetch data from Cricsheet.org ------------------------

wbbl07_bbb <- fetch_cricsheet_bbb("wbbl") %>%
  filter(season == "2021/22")
wbbl07_match_info <- fetch_cricsheet_match_info("wbbl") %>%
  filter(season == "2021/22")

# Do a little tidying up first -------------------------------------------------

wbbl07_match_info %>%
  select(
    match_id, winner, winner_runs, winner_wickets, method, outcome,
    eliminator) %>%
  mutate(
    match_id = factor(match_id)) -> wbbl07_match_info_tidy

wbbl07_bbb %>%
  tidyr::separate(ball, c("over_num_extracted", "ball_num")) %>%
  mutate(
    match_id = factor(match_id),
    over_num_numeric = as.numeric(over_num_extracted) + 1,
    runs_scored = runs_off_bat + extras,
    ball_num = as.numeric(ball_num),
    wicket_type = case_when(
      wicket_type == "" ~ NA_character_,
      TRUE              ~ wicket_type)) %>%
  group_by(match_id, innings) %>%
  mutate(
    temp_var = 1,
    balls_cumulative = cumsum(temp_var),
    runs_cumulative = cumsum(runs_scored),
    runs_total = max(runs_cumulative)) %>%
  ungroup() %>%
  select(-temp_var) %>%
  left_join(., wbbl07_match_info_tidy) %>%
  mutate(
    outcome_batting_team = case_when(
      outcome %in% c("no result", "tie") ~ as.character(outcome),
      winner == batting_team             ~ "won",
      TRUE                               ~ "lost"),
    outcome_bowling_team = case_when(
      outcome %in% c("no result", "tie") ~ as.character(outcome),
      winner == bowling_team             ~ "won",
      TRUE                               ~ "lost")
    ) -> wbbl07_bbb_tidy

```

# Team strike rate after wickets

One tactical consideration that makes the shorter formats so fascinating is how to go about scoring runs after you've lost a wicket.

With only 120 regulation balls bowled in a 20-over match:

- How many balls do you take to get your eye in?
- When do you decide to push on and swing the bat?
- How does this differ with "game state", when the wickets and overs remaining equation changes?

One way we can look at this is to examine patterns in team strike rate (runs scored for balls faced) across the course of an innings.

```{r team_strike_rate}

wbbl07_bbb_tidy %>%
  # Exclude matches that ended with a Super Over
  # and abandoned matches
  filter(!outcome_batting_team %in% c("tie", "no result")) %>%
  group_by(match_id, innings) %>%
  mutate(
    rolling_strike_rate = round(
      runs_cumulative / balls_cumulative * 100, 1),
    wicket_ball_num = case_when(
      !is.na(wicket_type) ~ balls_cumulative,
      TRUE                ~ NA_real_),
    wicket_strike_rate = case_when(
      !is.na(wicket_type) ~ rolling_strike_rate,
      TRUE                ~ NA_real_),
    innings_description = case_when(
      innings == 1 ~ "Setting",
      innings == 2 ~ "Chasing"),
    match_details = glue::glue(
      "{innings_description} vs. {bowling_team}\n({start_date})")) %>%
  arrange(match_id, innings, balls_cumulative) %>%
  mutate(
    match_details = factor(
      match_details, levels = unique(match_details))) -> team_strike_rate

```

## Plot: Renegades

Let's take a look at the Melbourne Renegades, who currently sit at the top of the standings (as of Nov 6 2021):

```{r add_google_fonts}

font_add_google("Roboto Condensed", "roboto_con")
font_add_google("Staatliches", "staat")

```

```{r plot_team_strike_rate_renegades}

# Filter to Renegades' innings only
team_strike_rate %>%
  filter(
    str_detect(batting_team, "Renegades")
    ) -> team_strike_rate_renegades

# Build plot
p <- ggplot(
  team_strike_rate_renegades,
  aes(x = balls_cumulative, y = rolling_strike_rate))
p <- p + facet_wrap(~ match_details, ncol = 4)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team))
p <- p + geom_point(
  aes(x = team_strike_rate_renegades$wicket_ball_num,
      y = team_strike_rate_renegades$wicket_strike_rate),
      colour = "red", size = 3, alpha = 0.75)
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank())

```

## Plot: Sydney Thunder

And now let's plot the same variables but for the Sydney Thunder, who have had a few losses to start their title defense and are currently at the bottom of the standings on net run rate:

```{r plot_team_strike_rate_thunder}

# Filter to Thunder innings only
team_strike_rate %>%
  filter(
    str_detect(batting_team, "Thunder")
    ) -> team_strike_rate_thunder

# Build plot
p <- ggplot(
  team_strike_rate_thunder,
  aes(x = balls_cumulative, y = rolling_strike_rate))
p <- p + facet_wrap(~ match_details, ncol = 4)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team))
p <- p + geom_point(
  aes(x = team_strike_rate_thunder$wicket_ball_num,
      y = team_strike_rate_thunder$wicket_strike_rate),
      colour = "red", size = 3, alpha = 0.75)
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank())

```

# Plot: Team strike rate from winning vs. losing teams

```{r plot_team_strike_rate_all}

# Build plot
p <- ggplot(
  team_strike_rate,
  aes(x = balls_cumulative, y = rolling_strike_rate, group = match_id))
p <- p + facet_grid(innings ~ outcome_batting_team)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team))
p <- p + geom_smooth(aes(group = outcome_batting_team), method = "loess")
#p <- p + geom_point(
#  aes(x = team_strike_rate$wicket_ball_num,
#      y = team_strike_rate$wicket_strike_rate),
#      colour = "red", size = 3, alpha = 0.75)
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank())

```

# Plot: Team strike rate by team

```{r plot_team_strike_rate_by_team}

# Build plot
p <- ggplot(
  team_strike_rate,
  aes(x = balls_cumulative, y = rolling_strike_rate, group = match_id))
p <- p + facet_grid(outcome_batting_team ~ batting_team)
p <- p + geom_hline(
  yintercept = 100, linetype = "dashed", colour = "#CCCCCC")
p <- p + geom_line(aes(colour = outcome_batting_team))
p <- p + geom_smooth(aes(group = outcome_batting_team), method = "loess")
#p <- p + geom_point(
#  aes(x = team_strike_rate$wicket_ball_num,
#      y = team_strike_rate$wicket_strike_rate),
#      colour = "red", size = 3, alpha = 0.75)
p <- p + scale_x_continuous(
  breaks = seq(0, 120, by = 30))
p <- p + coord_cartesian(ylim = c(0, 200))
p <- p + theme_minimal()
p <- p + theme(
  panel.grid.major.y = element_blank(),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank())

```