---
title: 'WNBL advanced stats: Effective field goal percentage'
author: Jacquie Tran
date: '2021-12-08'
slug: wnbl-efg-pct
categories:
  - On-field analyses
tags:
  - women's sport
  - basketball
  - sports analytics
  - r stats
draft: yes
---

# Introduction

Let's take a look at **effective field goal percentages** in the WNBL, prompted by this tweet from Paul Flynn (Assistant Coach, Melbourne Boomers WNBL):

<center><blockquote class="twitter-tweet"><p lang="en" dir="ltr">Would be keen to view any data on points per possession, pace and correlating eFG% over the same period.</p>&mdash; Paul Flynn (@paulflynn611) <a href="https://twitter.com/paulflynn611/status/1466018173375836161?ref_src=twsrc%5Etfw">December 1, 2021</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script></center>

# Getting started

## Load R libraries

Load up the libraries we need for this analysis.

```{r load_libraries, message = FALSE, warning = FALSE}

# Load libraries
library(dplyr)      # for tidying data
library(tidyr)      # for tidying data
library(gt)         # for making data tables
library(showtext)   # for loading custom fonts for plots
library(ggplot2)    # for plotting
library(ggbeeswarm) # for plotting
library(ggtext)     # for formatting text in ggplot2 objects
library(corrr)      # for running correlations

```

## Read data in

We're going to use box scores data recorded at the team-level.

```{r read_data_in}

box_scores <- readRDS(url("https://github.com/jacquietran/wnblr_data/raw/main/data/box_scores.rds"))

```

## Tidying data

**Effective field goal percentage** is a metric for evaluating the effectiveness of 2-pt and 3-pt shots.

Per [NBAStuffer.com](https://www.nbastuffer.com/analytics101/effective-field-goal-percentage-efg/), effective field goal percentage is calculated as follows:

> Effective field goal percentage = ((FGM) + 0.5*(3PM))/FGA

(Where: FGM = Field goals made; 3PM = Three pointers made; FGA = Field goals attempted.)

```{r tidy_data, echo = FALSE, cache = TRUE}

# Tidy up, including calculating new variables ---------------------------------

box_scores %>%
  # Exclude games from 2021 season that is underway
  filter(season != 2021) %>%
  # Tidy up teams that changed names
  mutate(
    team_name_adj = case_when(
      team_name == "West Coast Waves"  ~ "Perth Lynx",
      team_name == "Dandenong Rangers" ~ "Southside Flyers",
      TRUE                             ~ team_name),
    team_name_opp_adj = case_when(
      team_name_opp == "West Coast Waves"  ~ "Perth Lynx",
      team_name_opp == "Dandenong Rangers" ~ "Southside Flyers",
      TRUE                                 ~ team_name_opp)) %>%
  # Focus on a selection of variables
  select(
    season, page_id, team_name_adj, team_name_opp_adj, points_total,
    points_total_opp, team_result, contains("field_goals"),
    contains("three_pointers"), contains("two_pointers"),
    contains("free_throws"), points_per_possession, possessions,
    possessions_opp, pace, efg_pct, efg_pct_opp) -> efg_tidy

```

# Analyses

## eFG% in wins and losses throughout the seasons

```{r}

# Summarise efg_pct for wins and losses, across the seasons --------------------

efg_tidy %>%
  group_by(season, team_name_adj, team_result) %>%
  summarise(
    efg_pct_mean = mean(efg_pct), .groups = "keep") %>%
  ungroup() -> efg_by_result_per_season

```

```{r}

p <- ggplot(
  efg_by_result_per_season,
  aes(x = season, y = efg_pct_mean, fill = team_result))
p <- p + facet_wrap(~team_name_adj)
p <- p + geom_bar(stat = "identity", position = "dodge")

```

## Team eFG% in 2020

```{r}

efg_tidy %>%
  filter(season == 2020) %>%
  mutate(
    efg_pct_relative = efg_pct - efg_pct_opp) -> efg_tidy_2020

p <- ggplot(
  efg_tidy_2020, aes(x = efg_pct_relative, y = team_name_opp_adj,
                     colour = team_result))
p <- p + facet_wrap(~team_name_adj)
p <- p + geom_point()

# Effective field goal percentage point difference
p

```

## Points per possession averages per team per season

```{r}

efg_tidy %>%
  group_by(season, team_name_adj) %>%
  summarise(
    points_per_possession_mean = mean(points_per_possession)) %>%
  ungroup() %>%
  arrange(season, desc(points_per_possession_mean)) %>%
  tidyr::pivot_wider(
    id_cols = team_name_adj,
    names_from = season,
    values_from = points_per_possession_mean)

```

# Team pace



```{r}

efg_tidy %>%
  filter(team_name_adj != "SEQ Stars") %>%
  group_by(season, team_name_adj, team_result) %>%
  summarise(pace_mean = mean(pace)) %>%
  ungroup() -> pace_tidy

```

```{r}

p <- ggplot(pace_tidy, aes(x = season, y = pace_mean, colour = team_result))
p <- p + facet_wrap(~team_name_adj)
p <- p + geom_line()

# Team average pace across seasons, in wins and losses
p

```


## Correlating points per possession and eFG%

```{r, message = FALSE}

efg_tidy %>%
  select(points_per_possession, efg_pct) %>%
  correlate(method = "pearson", use = "pairwise.complete.obs") %>%
  filter(term == "points_per_possession") %>%
  select(efg_pct) %>%
  mutate(efg_pct = round(efg_pct, 2)) %>%
  pull() -> cor_ppp_vs_efg_pct

```

```{r}

p <- ggplot(
  efg_tidy, aes(x = points_per_possession, y = efg_pct))
p <- p + facet_wrap(~team_name)
p <- p + geom_point(size = 3, alpha = 0.5)
p <- p + geom_smooth(method = "lm")

```


## Correlating pace and eFG%

```{r}


```


## Effective field goal percentage and days rest



```{r, echo = FALSE}

showtext_auto(FALSE)

```
