---
title: Exploring music data with {spotifyr}
author: Jacquie Tran
date: '2020-09-12'
slug: exploring-music-data-with-spotifyr
categories:
  - Music
tags:
  - music
  - spotifyr
draft: yes
---

Load up the necessary R packages.

```{r load_libraries, message = FALSE}

library(spotifyr)
library(dplyr)
library(tidyr)
library(ggplot2)

```

Retrieve data using the Spotify API.

```{r import_data}

# API authentication details already stored as environment variables

# Retrieve access token
access_token <- get_spotify_access_token()

# Retrieve audio features of selected playlists
# that I listen to while running
playlists_for_running <- get_playlist_audio_features(
  username = "jacquietran",
  playlist_uris = c("51vGYzueZGouaZm1ZqcKDP", # Aug 2020
                    "0KZbG6sTuMvBMNVc6g0Xbt", # Summer runnin'
                    "66mV0cj4SIGF89AfyP2UKM"), # Run things
    authorization = access_token)

```

Tidy up the data we've retrieved to focus on only the variables we need.

```{r tidy_data}

playlists_for_running %>%
  select(playlist_id, playlist_name, danceability, energy, key, key_name, mode,
         mode_name, key_mode, loudness, speechiness, acousticness,
         instrumentalness, liveness, valence, tempo, track.id, time_signature,
         track.artists, track.name) %>%
  unnest(track.artists) %>%
  select(-href, -id, -type, -uri, -external_urls.spotify) %>%
  rename(track_artists = name) -> playlists_for_running_tidy

```

Create a subset of the tidy data that focuses on the audio feature variables, and reshapes it into long format for plotting.

```{r audio_features}

# Subset to audio features -----------------------------------------------------

playlists_for_running_tidy %>%
  select(-contains("key"), -contains("mode"), -time_signature) %>%
  pivot_longer(
    cols = c("danceability", "energy", "loudness", "speechiness",
             "acousticness", "instrumentalness", "liveness", "valence",
             "tempo"),
    names_to = "audio_feature_name",
    values_to = "audio_feature_value") -> playlists_audio_features_long

```

Build a plot to compare the audio features of the 3 playlists I usually choose from when I'm going for a run.

```{r plot}

# Build plot -------------------------------------------------------------------

p <- ggplot(
  playlists_audio_features_long,
  aes(x = audio_feature_value, y = playlist_name,
      group = playlist_name, colour = audio_feature_name))
p <- p + facet_wrap(~ audio_feature_name, nrow = 3, scales = "free_x")
p <- p + geom_boxplot(
  width = 0.75, size = 1, outlier.size = -1)
p <- p + geom_jitter(
  height = 0.2, alpha = 0.25)
p <- p + theme(
  legend.position = "none",
  panel.spacing = unit("1", "lines"))

# Display plot -----------------------------------------------------------------
p

```