---
title: Advanced stats for the WNBL
author: Jacquie Tran
date: '2021-12-02'
slug: []
categories:
  - On-field analyses
tags:
  - wnbl
  - r stats
  - sports analytics
  - sport science
draft: yes
---

# Introduction

Dean Oliver's Four Factors: https://squared2020.com/2017/09/05/introduction-to-olivers-four-factors/

{wnblr} package includes live stats recorded from WNBL games, but doesn't include "advanced" stats like effective field goal percentage, points per possession, etc.

Paul Flynn (Assistant Coach, Melbourne Boomers) asked to see some specific advanced stats from WNBL play:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Would be keen to view any data on points per possession, pace and correlating eFG% over the same period.</p>&mdash; Paul Flynn (@paulflynn611) <a href="https://twitter.com/paulflynn611/status/1466018173375836161?ref_src=twsrc%5Etfw">December 1, 2021</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script> 

# Getting started

Load up the libraries we need. The {wnblr} package provides access to built-in data sets with game stats.

```{r}

# remotes::install_github("jacquietran/wnblr")
library(wnblr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggbeeswarm)

```

# Tidying data

- Points per possession: https://www.nbastuffer.com/analytics101/points-per-possession-ppp/
     - Possession: https://www.nbastuffer.com/analytics101/possession/
- Pace: https://www.nbastuffer.com/analytics101/pace/
- eFG%: https://www.nbastuffer.com/analytics101/effective-field-goal-percentage-efg/

Firstly, we need to calculate number of possessions.

```{r poss, eval = FALSE}

# Go through calculations step by step, but evaluate in one dplyr chain

# Via NBAStuffer.com: Basic Possession Formula=0.96*[(Field Goal Attempts)+(Turnovers)+0.44*(Free Throw Attempts)-(Offensive Rebounds)]

box_scores %>%
  mutate(
    num_possessions = 0.96*(field_goals_attempted + turnovers + (0.44*free_throws_attempted) - rebounds_offensive))

```

```{r points_per_poss, eval = FALSE}

# Via NBAStuffer.com: Points Per Possession =PTS/(FGA+0.44*FTA+TO)

```

```{r pace, eval = FALSE}

# Via NBAStuffer.com: Pace Calculation Formula=[240/(Team Minutes)]*(Possessionteam+Possessionopponent)/2

```

```{r efg_pct, eval = FALSE}

# Via NBAStuffer.com: Effective Field Goal Percentage Formula=[(All Field Goals Made) + 0.5*(3P Field Goals Made)]/(All Field Goal Attempts)

```

```{r, echo = FALSE}

box_scores %>%
  mutate(
    # Number of possessions - calculate in separate columns
    # for home and away (so we can later have two columns:
    # one for number of possessions of `team_name` and one for `team_name_opp`)
    num_possessions_home_team = case_when(
      home_away_flag == "home" ~ round(
        (0.96*(field_goals_attempted + turnovers + (0.44*free_throws_attempted) - rebounds_offensive)), 1),
      TRUE                     ~ NA_real_),
    num_possessions_away_team = case_when(
      home_away_flag == "away" ~ round(
        (0.96*(field_goals_attempted + turnovers + (0.44*free_throws_attempted) - rebounds_offensive)), 1),
      TRUE                     ~ NA_real_)) %>%
  group_by(page_id) %>%
  fill(num_possessions_home_team) %>%
  fill(num_possessions_away_team, .direction = "up") %>%
  ungroup() %>%
  mutate(
    num_possessions = case_when(
      home_away_flag == "home" ~ num_possessions_home_team,
      home_away_flag == "away" ~ num_possessions_away_team),
    num_possessions_opp = case_when(
      home_away_flag == "home" ~ num_possessions_away_team,
      home_away_flag == "away" ~ num_possessions_home_team),
    # Points per possession
    points_per_possession = round((points_total / (field_goals_attempted + (0.44 * free_throws_attempted) + turnovers)), 3)) %>%
    # This next bit of code is a workaround to tidy up the janky minutes values
    # that I need to properly address in data scraping and cleaning for {wnblr}
  filter(!page_id %in% c(1087574, 803093)) %>%
  separate(
      minutes, into = c("min_extracted", "s_extracted"),
      sep = "\\:", remove = FALSE) %>%
  mutate(
    min_extracted = as.numeric(min_extracted),
    min_extracted = case_when(
      min_extracted < 210   ~ "200:00",
      min_extracted > 220 &
        min_extracted < 230 ~ "225:00",
      TRUE                  ~ "250:00"),
    minutes_numeric = case_when(
      min_extracted == "200:00" ~ 200,
      min_extracted == "225:00" ~ 225,
      min_extracted == "250:00" ~ 250)) %>%
  select(-minutes, -s_extracted) %>%
  rename(minutes = min_extracted) %>%
  mutate(
    # Pace per 40 min = [200/(Team Minutes)]*(Possessionteam+Possessionopponent)/2
    # Adjusted to per 40 min (for WNBL) instead of per 48 min (for NBA)
    pace = round(
      (200 / minutes_numeric) * ((num_possessions + num_possessions_opp)/2), 1),
    # eFG%
    efg_pct = round((((field_goals_made + (0.5 * three_pointers_made)) / field_goals_attempted) * 100), 1)) %>%
  # Exclude extraneous variables
  select(
    -num_possessions_home_team, -num_possessions_away_team
    ) -> box_scores_advanced

```

## Taking a closer look - 2020 season

```{r subset_to_2020}

# Focus on a few select variables ----------------------------------------------

box_scores_advanced %>%
  filter(season == 2020) %>%
  select(
    page_id, team_name, team_name_opp, points_total, points_total_opp,
    team_result, minutes_numeric, contains("field_goals"),
    contains("three_pointers"), contains("two_pointers"),
    contains("free_throws"), contains("num_possessions"), points_per_possession,
    pace, efg_pct
    ) -> selected_stats_2020

```

High / low scoring rate / efficiency (high points per possession)

```{r}

# Wrangle summary stats
selected_stats_2020 %>%
  group_by(team_name) %>%
  summarise(
    points_per_possession_mean = round(
      mean(points_per_possession), 3),
    points_per_possession_sd = round(
      sd(points_per_possession), 3),
    points_per_possession_sd_lower = points_per_possession_mean - points_per_possession_sd,
    points_per_possession_sd_higher = points_per_possession_mean + points_per_possession_sd) %>%
  ungroup() %>%
  mutate(
    team_name = factor(
      team_name, levels = c(
        "Bendigo Spirit",
        "Sydney Uni Flames",
        "Perth Lynx",
        "Adelaide Lightning",
        "UC Capitals",
        "Melbourne Boomers",
        "Townsville Fire",
        "Southside Flyers"))
    ) -> points_per_possession_summarised

# Wrangle game stats to order factor levels
selected_stats_2020 %>%
  mutate(
    team_name = factor(
      team_name, levels = c(
        "Bendigo Spirit",
        "Sydney Uni Flames",
        "Perth Lynx",
        "Adelaide Lightning",
        "UC Capitals",
        "Melbourne Boomers",
        "Townsville Fire",
        "Southside Flyers"))) %>%
  # Merge with summary stats
  left_join(
    ., points_per_possession_summarised
    ) -> points_per_possession

# Build plot
p <- ggplot(
  points_per_possession,
  aes(x = points_per_possession, y = team_name, colour = team_name))
p <- p + geom_jitter(height = 0.2, alpha = 0.5, size = 6)
p <- p + geom_point(
  aes(x = points_per_possession_mean, y = team_name),
  colour = "#FFFFFF", size = 7)
p <- p + geom_linerange(
  aes(xmin = points_per_possession_sd_lower,
      xmax = points_per_possession_sd_higher),
  colour = "#FFFFFF", size = 1, linetype = "dotted")
p <- p + geom_text(
  aes(x = points_per_possession_mean, y = team_name,
      label = points_per_possession_mean),
  hjust = -0.25, vjust = -0.5, colour = "#FFFFFF", fontface = "bold", size = 6)
p <- p + ggdark::dark_mode()
p <- p + theme(
  legend.position = "none")

```

High / low pace teams?

```{r}

```


High / low effective field goal percentage (value for shots taken)

```{r}

```

Correlation between points per possession and eFG%

```{r}

```

Correlation between pace per 40 min and eFG%

```{r}

```
